### 2023-11-03 Code/460ProjectDraft.Rmd
R Markdown draft

---
title: "460ProjectDraft"
author: "James Santarpio"
date: "2023-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-11-03 Code/RmarkdownDraft.R

@@ -1,61 +1,0 @@
----
-title: "Draft Exploratory Analysis"
-author: "James Santarpio and Mardi Elings"
-date: "2023-11-03"
-output:
-  html_document:
-    df_print: paged
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-# Data Analytics Term Project
-## Project Desription 
-
-### Topic:
-The effect of drug decriminalization on overdose deaths and well-being.
-
-### Why it Matters 
-The study of drug decriminalization and overdose rates in Oregon has potential for significant policy and societal relevance. There are many debates on the benefits and drawbacks of drugs use and decriminalization has potential for policy implications involving economics, public health, criminal justice, and community well-being. 
-
-### Research Question and Motivation
-What is the effect of drug decriminalization on overdose deaths and well-being? A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths and the economic impacts on health care costs and overall well-being. We used the synthetic control method to understand how the policy change impacted Oregon by comparing what would have happened in th absence of the policy being implemented.  
-
-### Description of Data
-
-#### Overdose Mortality by County in the US
-The Overdose Mortality by County in the US is from the CDC Center for Health Statistics for the years 2020-2023. These provisional county-Level drug Overdose death counts are based on death records and is received  by the NCHS receives the death records from the state vital registration offices through the Vital Statistics Cooperative Program (VSCP). The provisional counts are reported by the county were the decedent resided, not necessarily where they died. 
-
-#### County Well-Being Rankings 
-The County Well-Being Rankings are for the years 2014-2023 and is collected County Health Rankings & Roadmaps CHR&R) is a program of the University of Wisconsin Population Health Institute. The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public heatlh rankings. 
-
-#### Overdose by Drug 
-The Overdose by drug dataset contains information on non-fatal overdose data by state for 2018-2023. Provided by the CDC, DOSE (Drug Overdose Surveillance and Epidemiology) collects health department reports in each state. 
-
-Other variables of interest include opioid and stimulant overdoses and the significance of the change from previous months in those overdoses.
-
-#### Violent Crime Data
-From the FBI crime data reporter database, the number of violent crimes in Oregon. 
-
-
-#### How are they related? 
[…]

### 2023-11-03 Code/460ProjectDraft.Rmd
Working Draft

@@
 title: "460ProjectDraft"
-author: "James Santarpio"
+author: "James Santarpio, Mardi"
 date: "2023-11-03"
@@
 ---
+Project Description
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+Topic:
+The effect of drug decriminilization on overdose deaths and well-being.
+
+Why it matters:
+The study of drug decriminalization and overdose rates in Oregon has potential for significant policy and societal relevance.
+There are many debates on the benefits and drawbacks of drugs use and decriminalization has potential for policy implications involving economics, public health, criminal justice, and community well-being.
+
+Research Question and Motivation:
+What is the effect of drug decriminalization on overdose deaths and well-being?
+A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths and the economic impacts on health care costs and overall well-being.
+We used the synthetic control method to understand how the policy change impacted Oregon by comparing what would have happened in th absence of the policy being implemented.
+
+Description of Data:
+
+Overdose Mortality by county in the US:
+The Overdose Mortality by County in the US is from the CDC Center for Health Statistics for the years 2020-2023.
+These provisional county-Level drug Overdose death counts are based on death records and is received  by the NCHS receives the death records from the state vital registration offices through the Vital Statistics Cooperative Program (VSCP).
+The provisional counts are reported by the county were the decedent resided, not necessarily where they died.
+
+County Well-Being Rankings:
+The County Well-Being Rankings are for the years 2014-2023 and is collected County Health Rankings & Roadmaps CHR&R) is a program of the University of Wisconsin Population Health Institute.
+The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public heatlh rankings.
+
+Overdose by Drug:
+The Overdose by drug dataset contains information on non-fatal overdose data by state for 2018-2023.
+Provided by the CDC, DOSE (Drug Overdose Surveillance and Epidemiology) collects health department reports in each state.
+
+Other variables of interest include opioid and stimulant overdoses and the significance of the change from previous months in those overdoses.
+
+Violent Crime Data
+From the FBI crime data reporter database, the number of violent crimes in Oregon.
+
+How are they related? 
+The Well-Being Ranks, Overdose Mortality by County contain county fips codes, county names and year in relation.
+The Overdose by Drug has a relationship of total non-fatal drug overdoses in each state.
+
+Data Processing:
+
+Findings:
+
+Limitations:
+
+Results:
+```{r setup}
+
 ```
 
-## R Markdown
+```{r setup}
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
 ```
 
-## Including Plots
+```{r setup}
 
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```
+
+```{r setup}
+
+```

### 2023-11-03 Code/Functions.R
Updating Main R File

@@ -1,8 +1,8 @@
 #GGplot histogram
-ggplot_hist <- function(df,x){
-  return(ggplot(df, aes(x = df$x))+
-           geom_histogram()
-  )
+ggplot_hist <- function(df,x){ #currently dont work, not sure why
+  library(ggplot2)
+  return(ggplot(df, aes(x = x))+
+           geom_histogram())
 }
 
 ggplot_bar <- function(df,x,y){
@@ -10,6 +10,9 @@ ggplot_bar <- function(df,x,y){
            geom_bar(stat = "identity"))
 }
 
+#Function to convert healthrank to yearly mean sums. 
+#The function removes the state total from the dataframe to accurately
+#Calculate yearly column means
 healthrank_to_yearly <- function(df){
   df <- df[!df$county == "year_state_total"]
   df <- df %>%

### 2023-11-03 Code/CleanginCode/MasterCleaningCode.R
population

@@ -480,6 +480,7 @@ colnames(county_health_ranked_measured2) <- sub("^_", "", colnames(county_health
 
 
 
+
 #******************* Addtional Measure Data All Sates***********************
 # Create a list for health rank data 
 health_rank_data_list2 <- list()
@@ -735,7 +736,7 @@ filtered2 <- combine2 |>
     x__drug_overdose_deaths,
     drug_overdose_mortality_rate,
     household_income,
-    population,
+    population
   )
 View(filtered2)
 
@@ -751,7 +752,8 @@ filtered3 <- combine3 |>
     infant_mortality_rate,
     x__drug_overdose_deaths,
     drug_overdose_mortality_rate,
-    median_household_income
+    median_household_income,
+    population
   )
 
 View(filtered1)
@@ -804,7 +806,6 @@ all_county_additional_ranks2 <- all_county_additional_ranks2 |>
 str(all_county_additional_ranks)
 
 View(all_county_additional_ranks2) # Check results
-
 
 #*************** Final merged data frame for health rankings **********
 

### 2023-11-03 Code/visualizations.R
Update visualizations.R

@@ -61,3 +61,15 @@ dataCA |>
   ggplot(aes(combined_overdose_mortality_rate, county)) +
   geom_line(aes(group = county)) +
   geom_point(size=2, aes(color = factor(year)))
+
+# Show deaths percapita for each state
+data2 <- data |>
+  mutate(od_deaths_percapita = combined_overdose_deaths/population) 
+
+data2 |>
+  filter(year %in% c(2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
+  ggplot(aes(od_deaths_percapita, y=..count.., fill = state)) +
+  scale_x_continuous(trans = "log10") + 
+  geom_density(alpha=0.3, position = "stack") + 
+  facet_grid(year ~ .) 
+

### 2023-11-03 Code/Functions.R
Main file

@@ -14,9 +14,8 @@ ggplot_bar <- function(df,x,y){
 #The function removes the state total from the dataframe to accurately
 #Calculate yearly column means
 healthrank_to_yearly <- function(df){
-  df <- df[!df$county == "year_state_total"]
+  df <- df[!df$county == "year_state_total",]
   df <- df %>%
-    select_if(function(col) !all(col = "year_state_total")) %>%
     group_by(year) %>%
     summarise(across(where(is.numeric), ~ mean(., na.rm = TRUE))) %>%
     select(-c("X", "fips"))

### 2023-11-03 Code/Main.R

@@ -1,115 +1,169 @@
 #ECNS 460 Final Project Draft and other stuff
+#Working Directory- Set to Project location
 
-#This will be main file and where to mess around in
-#Working directory is set to project
-
-#Load in functions (Currently none in Functions)
+#Load in functions 
 source("Code/Functions.R")
+
 library(tidyverse)
+library(ggplot2)
 
-#Current Data we have:
-#Oregon Violent Crime per year- https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/explorer/crime/crime-trend
-#Called OregonViolentCrim
+health_ranking <- read.csv("Data/Cleaned Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
 
+#Remove 2023- 2023 is a duplicate of 2022 for much of the data
+health_ranking <- health_ranking %>%
+  filter(year <= 2022) 
 
-#Overdose by drug for each state- https://www.cdc.gov/drugoverdose/nonfatal/dose/surveillance/dashboard/index.html#stateInfo
-#Called overdose_by_drug
-
-#Overdose deaths by county 2020 - now - https://www.cdc.gov/nchs/nvss/vsrr/prov-county-drug-overdose.htm
-#Called OverdoseDeathsByCounty2020-
-
-#Arrests Data 2020- https://www.oregon.gov/osp/pages/uniform-crime-reporting-data.aspx
-
-or_health_ranking <- read.csv("Data/Cleaned Data/CleanedStateHealthRankings/OregonCountyHealthRank2018_2023.csv")
+#DO we need these two any more?
 national_overdose_by_county <- read.csv("Data/Cleaned Data/NationalOverdoseDeathsByCounty_cleaned.csv")
-oregon_overdose_deaths_by_county <- read.csv("Data/Cleaned Data/OregonOverdoseDeathsByCounty2020_2023.csv")
+oregon_overdose_deaths_by_county <- 
+  national_overdose_by_county[national_overdose_by_county$ST_ABBREV == "OR",]
+
 violent_crime <- read.csv("Data/Cleaned Data/ViolentCrime.csv")
 overdose_by_drug <- read.csv("Data/Cleaned Data/OverDosebyDrug_cleaned.csv")
+
+#Not sure if this is necessary but here for now
 oregon_merged_data <- read.csv("Data/Cleaned Data/OregonMergedCountyData.csv")
 
-#Start with Oregon Violent Crime:
[…]

### 2023-11-03 Code/visualizations.R
Update visualizations.R

@@ -66,6 +66,7 @@ dataCA |>
 data2 <- data |>
   mutate(od_deaths_percapita = combined_overdose_deaths/population) 
 
+# All states and overdose death percapita for years 2016, 2020, 2023
 data2 |>
   filter(year %in% c(2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
   ggplot(aes(od_deaths_percapita, y=..count.., fill = state)) +
@@ -73,3 +74,13 @@ data2 |>
   geom_density(alpha=0.3, position = "stack") + 
   facet_grid(year ~ .) 
 
+# Scatter of years of potential life lost rate and overdose mortality rate.
+data2 %>%
+      filter(year %in% c(2020, 2021)) %>%
+      ggplot(aes(x = combined_overdose_mortality_rate, y = combined_ypll_rate, color = factor(year))) +
+      geom_point() +
+      theme_gray() +
+      labs(title = "Scatter Plot for 2020 and 2021", x = "Overdose Mortality Rate", y = "Years Potential Life Lost Rate/100,000 People") +
+      scale_color_manual(values = c("2020" = "blue", "2021" = "red"))
+    
+    

### 2023-11-03 Code/Main.R
Updated Main R FIle

@@ -69,64 +69,48 @@ or_yearly <- healthrank_to_yearly(or_health_ranking)
 ca_yearly <- healthrank_to_yearly(ca_health_ranking)
 wa_yearly <- healthrank_to_yearly(wa_health_ranking)
 
-ggplot()+
-  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths), color = "blue")+
-  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths), color = "red")+
-  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths), color = "green")
-
-plot(or_yearly$combined_overdose_deaths) 
-plot(ca_yearly$combined_overdose_deaths)
-plot(wa_yearly$combined_overdose_deaths)
-
-#Visually looks like no oregon may be increasing at a slower rate compared to wa/ca
-
-
-#Data is missing a good amount of info. It will be assumed this is missing at random. Data appears to be missing for each state the same way (wont affect synthetic control)
-
-
-#Create pre and post 2020 for before and after the law was enacted
-or_health_ranking_pre <- or_yearly %>%
-  filter(year <= 2020)
-or_health_ranking_post <- or_yearly %>%
-  filter(year > 2020)
-
-ca_health_ranking_pre <- ca_yearly %>%
-  filter(year <= 2020)
-ca_health_ranking_post <- ca_yearly %>%
-  filter(year > 2020)
-
-wa_health_ranking_pre <- wa_yearly %>%
-  filter(year <= 2020)
-wa_health_ranking_post <- wa_yearly %>%
-  filter(year > 2020)
-
 #lets look at combined overdose deaths
 ggplot()+
-  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths), color = "blue")+
-  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths), color = "red")+
-  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths), color = "green")
-  
+  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths, color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths, color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
+  scale_color_manual(name = "State",
+                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  theme_bw()
+
[…]

### 2023-11-03 Code/460ProjectDraft.Rmd
Update 460ProjectDraft.Rmd

@@
 
+For the county health rankings data, we originally had years 2014-2023, but after further inspection the way overdose deaths were collected for 2014-2015 was significantly different that for the years 2016-2023.
+We opted to drop those years since those data sets also did not include as many well-being measures.
+
 Findings:
@@
 Limitations:
+It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths.
+Since multiple factors can contribute to drug-related outcomes.
+There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health.
+There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon’s drug policies and may not be generalized to other regions or policies.
 

### 2023-11-04 Code/Main.R
Updated Main File

@@ -13,15 +13,9 @@ health_ranking <- read.csv("Data/Cleaned Data/CleanedStateHealthRankings/MergedA
 health_ranking <- health_ranking %>%
   filter(year <= 2022) 
 
-#DO we need these two any more?
-national_overdose_by_county <- read.csv("Data/Cleaned Data/NationalOverdoseDeathsByCounty_cleaned.csv")
-oregon_overdose_deaths_by_county <- 
-  national_overdose_by_county[national_overdose_by_county$ST_ABBREV == "OR",]
-
 violent_crime <- read.csv("Data/Cleaned Data/ViolentCrime.csv")
 overdose_by_drug <- read.csv("Data/Cleaned Data/OverDosebyDrug_cleaned.csv")
 
-#Not sure if this is necessary but here for now
 oregon_merged_data <- read.csv("Data/Cleaned Data/OregonMergedCountyData.csv")
 
 #Oregon Violent Crime:
@@ -35,29 +29,24 @@ ggplot(violent_crime, aes(x = Year, y = USA))+
 USOverdose_by_drug <- overdose_by_drug[overdose_by_drug$st_abbrev == "US",]
 OROverdose_by_drug <- overdose_by_drug[overdose_by_drug$st_abbrev == "OR",]
 
+#This data will be used to compare with CA, OR, and WA overdose death totals.
+
+''' Will come back to this:
 library(binsreg)
 binsreg(OROverdose_by_drug$all_percentagechange, OROverdose_by_drug$start_year)
 ggplot(OROverdose_by_drug, aes(x = start_year, y = all_percentagechange))+
   geom_point()
-#*****************FIX Y AXIS******************
+
 ggplot(USOverdose_by_drug, aes(x = start_year, y = all_percentagechange))+
   geom_point()
-#******************VSUALIZE BETTER*****************
+'''
 #What about the is significant column? This column represents a statistically significant, either
 #an increase or decrease
 
 table(USOverdose_by_drug$all_issignificant)
 table(OROverdose_by_drug$all_issignificant)
 
-#Oregon has lower is significant values relative to the U.S.However, this includes decreases as well.
-
-#USA:
-#Lotta N/A values
-
-sum(is.na(national_overdose_by_county))/nrow(national_overdose_by_county)
-#46% of data is missing. This is significant.
-#This data will be used to compare with CA, OR, and WA overdose death totals.
-
+#Oregon has lower is significant values relative to the U.S.
 
[…]

### 2023-11-04 Code/Main.R
Updated readme and main R file

@@ -40,6 +40,7 @@ ggplot(OROverdose_by_drug, aes(x = start_year, y = all_percentagechange))+
 ggplot(USOverdose_by_drug, aes(x = start_year, y = all_percentagechange))+
   geom_point()
 '''
+
 #What about the is significant column? This column represents a statistically significant, either
 #an increase or decrease
 
@@ -144,6 +145,6 @@ path.plot(synth.res = synth.out,
           Legend.position = c("topleft")
 )
 
-#Our synthetic control analysis seems innacurate. The synthetic Oregon does not follow actual Oregon well pre treatment.
+#Our synthetic control analysis may be innacurate. The synthetic Oregon does not follow actual Oregon well pre treatment.
 
 #The main takeaway is that synthetic Oregon appears to increase faster than actual Oregon in overdose deaths post 2020.

### 2023-11-04 Code/Main.R
Updated main R file

@@ -148,3 +148,4 @@ path.plot(synth.res = synth.out,
 #Our synthetic control analysis may be innacurate. The synthetic Oregon does not follow actual Oregon well pre treatment.
 
 #The main takeaway is that synthetic Oregon appears to increase faster than actual Oregon in overdose deaths post 2020.
+#This suggests decriminalizing drugs helped slow the increase in drug overdose deaths.

### 2023-11-04 Code/460ProjectDraft.Rmd
Updated rmd file

@@
 ---
+
 Project Description
@@
 A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths and the economic impacts on health care costs and overall well-being.
-We used the synthetic control method to understand how the policy change impacted Oregon by comparing what would have happened in th absence of the policy being implemented.
+We used the synthetic control method to understand how the policy change impacted Oregon by comparing what would have happened in the absence of the policy being implemented.
 
@@
 County Well-Being Rankings:
-The County Well-Being Rankings are for the years 2014-2023 and is collected County Health Rankings & Roadmaps CHR&R) is a program of the University of Wisconsin Population Health Institute.
+The County Well-Being Rankings are for the years 2014-2023 and is collected by County Health Rankings & Roadmaps CHR&R, a program of the University of Wisconsin Population Health Institute.
 The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public heatlh rankings.
@@
 Data Processing:
-
 For the county health rankings data, we originally had years 2014-2023, but after further inspection the way overdose deaths were collected for 2014-2015 was significantly different that for the years 2016-2023.
@@
 
+The data collected was generally in clean condition.
+SOme column name changes were done, and general data manipulation was completed.
+For example, health rankings came from yearly files- these had to be merged together.
+There was generally little missing data.
+Violent crime and overdose by drugs contained zero missing values.
+However, a key data set used, health rankings, contained 13 percent N/A values.
+These appeared generally randomly scatter, and we will assume these are missing at random and will be ignored.
+
 Findings:
+Initial findings include general decrease in well being, likely due to covid19 (mental health days, pct fair or poor health days, etc).
+Additionally, drug overdose deaths have been increasing, especially since 2020.
+However, it appears that Oregon's increase may be slower relative to Washington and Oregon after decriminilzaing drugs.
 
@@
 There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon’s drug policies and may not be generalized to other regions or policies.
+There is also limited data post treatment period (2 years).
+Additionally, the synthetic Oregon may not be a good fit.
+This wil be further analyzed throughout the semester.
+As well as continuing to fully understand the statistical basis for synthetic control better.
 
 Results:
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
-
-```{r setup}
-
-```
+Decriminilizing drugs in Oregon has slowed the increase in drug overdose deaths.

### 2023-11-27 Code/Main.R
DID in progress

@@ -41,8 +41,7 @@ ggplot(USOverdose_by_drug, aes(x = start_year, y = all_percentagechange))+
   geom_point()
 '''
 
-#What about the is significant column? This column represents a statistically significant, either
-#an increase or decrease
+#What about the is significant column? This column represents a statistically significant increase or decrease
 
 table(USOverdose_by_drug$all_issignificant)
 table(OROverdose_by_drug$all_issignificant)
@@ -66,6 +65,7 @@ ggplot()+
   geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
   scale_color_manual(name = "State",
                      values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  labs(y = "Overdose Deaths") +
   theme_bw()
 
 #Now do a log transformation to better show percent differences due to population size differences
@@ -149,3 +149,42 @@ path.plot(synth.res = synth.out,
 
 #The main takeaway is that synthetic Oregon appears to increase faster than actual Oregon in overdose deaths post 2020.
 #This suggests decriminalizing drugs helped slow the increase in drug overdose deaths.
+
+#Difference in difference analysis
+#Oregon, Washington, and California combined overdose deaths
+
+#Visualize again
+
+ggplot()+
+  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths, color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths, color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
+  scale_color_manual(name = "State",
+                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  labs(y = "Overdose Deaths") +
+  theme_bw()
+
+#Now do a log transformation to better show percent differences due to population size differences
+ggplot()+
+  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
+  scale_color_manual(name = "State",
[…]

### 2023-11-28 Code/Main.R
DID

@@ -184,7 +184,38 @@ ggplot()+
 
 #Treatment * Post is a dummy variable indicating whether the outcome was observed in the treatment group AND it was observed after the intervention (=1), or any other case (=0).
 
+
 #Create data frame contining oregon, washington, and Cali
 
-overdose_deaths <- as.data.frame(c(or_yearly[,1], or_yearly[,4], ca_yearly[,4], wa_yearly[,4]))
-colnames(overdose_deaths) <- c("Year", "Oregon_Overdoses", "California_Overdoses", "Washington_Overdoses")
+overdose_deaths <- rbind(wa_yearly[,c(1,4)], rbind(or_yearly[,c(1,4)], ca_yearly[,c(1,4)]))
+overdose_deaths$Country <- 0; overdose_deaths$Treatment <- 0
+
+overdose_deaths$Country[1:7] <- 1; overdose_deaths$Country[8:14] <- 2; overdose_deaths$Country[15:21] <- 3;
+
+#Washington-1, Oregon-2, California-3
+
+overdose_deaths$Treatment[overdose_deaths$year >= 2021] <- 1
+#Now can do DID analysis
+
+#Y = B0+B1*Group + B2*Treatment + B3*Group:Treatment + E
+
+did_model <- lm(combined_overdose_deaths ~ Country + Treatment + Country:Treatment, data = overdose_deaths)
+summary(did_model)
+
+
+
+overdose_deaths <- rbind(or_yearly[,c(1,4)], ca_yearly[,c(1,4)])
+overdose_deaths$Country <- 0; overdose_deaths$Treatment <- 0
+
+overdose_deaths$Country[1:7] <- 0; overdose_deaths$Country[8:14] <- 1; 
+
+#Washington-1, Oregon-2, California-3
+
+overdose_deaths$Treatment[overdose_deaths$year >= 2021] <- 1
+#Now can do DID analysis
+
+#Y = B0+B1*Group + B2*Treatment + B3*Group:Treatment + E
+
+did_model <- lm(combined_overdose_deaths ~ Country + Treatment + Country:Treatment, data = overdose_deaths)
+summary(did_model)
+

### 2023-11-28 Code/Main.R
DID

@@ -205,17 +205,16 @@ summary(did_model)
 
 
 overdose_deaths <- rbind(or_yearly[,c(1,4)], ca_yearly[,c(1,4)])
-overdose_deaths$Country <- 0; overdose_deaths$Treatment <- 0
+overdose_deaths$Treatment <- 0; overdose_deaths$Post <- 0
 
-overdose_deaths$Country[1:7] <- 0; overdose_deaths$Country[8:14] <- 1; 
+overdose_deaths$Treatment[1:7] <- 0; overdose_deaths$Treatment[8:14] <- 1; 
 
 #Washington-1, Oregon-2, California-3
 
-overdose_deaths$Treatment[overdose_deaths$year >= 2021] <- 1
+overdose_deaths$Post[overdose_deaths$year >= 2021] <- 1
 #Now can do DID analysis
 
-#Y = B0+B1*Group + B2*Treatment + B3*Group:Treatment + E
+#Y = B0+B1*Treatment + B2*Post + B3*Treatment:Post + E
 
-did_model <- lm(combined_overdose_deaths ~ Country + Treatment + Country:Treatment, data = overdose_deaths)
+did_model <- lm(combined_overdose_deaths ~ Treatment + Post + Treatment:Post, data = overdose_deaths)
 summary(did_model)
-

### 2023-11-29 Code/Main.R
county diff in diff

@@ -218,3 +218,17 @@ overdose_deaths$Post[overdose_deaths$year >= 2021] <- 1
 
 did_model <- lm(combined_overdose_deaths ~ Treatment + Post + Treatment:Post, data = overdose_deaths)
 summary(did_model)
+
+
+#******** Difference in Difference for county level data ****************
+# Filter out rows where 'county' is equal to "year_state_total" from the data
+health_ranking <- data |>
+  filter(county != "year_state_total")
+
+# Create a DiD treatment variable
+did_data <- health_ranking |>
+  mutate(treatment = ifelse(state == "Oregon" & year >= 2021, 1, 0))
+
+# Fit the DiD model using year and county fixed effects and population as a covariate
+did_model_county <- lm(combined_overdose_deaths ~ treatment + factor(county) + factor(year) + treatment:factor(year) + population, data = did_data)
+summary(did_model_county)

### 2023-11-29 Code/CleanginCode/MasterCleaningCode.R
Update MasterCleaningCode.R

@@ -3,10 +3,14 @@ library(dplyr)
 library(skimr)
 library(summarytools)
 
+# Set working directory 
+working_directory <- "/Users/mardielings/Desktop/Data Analytics/project/"
+setwd(working_directory)
+
 #************* Cleaning Oregon Violent Crime *********
 
 # Load Oregon  violent crime data 
-OregonViolentCrime <- "https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/OregonViolentCrime.csv"
+OregonViolentCrime <- "raw_data/OregonViolentCrime.csv"
 
 oregon_violent_crime <- read.csv(OregonViolentCrime, header = TRUE)
 # Rename column names
@@ -22,7 +26,8 @@ print(oregon_violent_crime)
 #********** Clean Overdose by County *****************
 
 # Load overdose by county 2020
-OverdoseByCounty2020 <- "https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/OverdoseDeathsByCounty2020-.csv"
+OverdoseByCounty2020 <- "raw_data/OverdoseByCount2020-.csv"
+
 overdose_by_county_2020 <- read.csv(OverdoseByCounty2020)
 
 # Standardize and rename column names
@@ -46,41 +51,17 @@ print(oregon_overdose_by_county) # Check results
 #write.csv(oregon_overdose_by_county, file = orod_file_path, row.names = FALSE)
 
 
-# load overdose by drug. sheet = 2 from excel workbook
-OverdosebyDrug <- "https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Overdose_by_drug.csv"
-overdose_by_drug <- read.csv(OverdosebyDrug)
 
-# remove rows with NA's and replace headers 
-overdose_by_drug <- overdose_by_drug |>
-  slice(4:n())
-colnames(overdose_by_drug) <- NULL
-
-# use first row to set column names and standardize col names
-colnames(overdose_by_drug) <- as.character(unlist(overdose_by_drug[1, ]))
-overdose_by_drug <- overdose_by_drug[-1, ]
-overdose_by_drug <- overdose_by_drug |>
-  rename_all(~str_replace(., "([A-Z])", "_\\1")) # places "_" before capital letter
-colnames(overdose_by_drug) <- tolower(colnames(overdose_by_drug)) # removes capitals from column headers
[…]

### 2023-11-29 Code/Draft1Rmd.Rmd
Add files via upload

---
title: "Exploratory Analysis"
author: "James Santarpio and Mardi Elings"
date: "2023-11-30"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggplot2)

url <- c("https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Cleaned%20Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
# Read csv
data <- read.csv(url)

# Remove 2023- 2023 is a duplicate of 2022 for much of the data
health_ranking <- data %>%
  filter(year <= 2023) %>%
  filter(! (county %in% c("year_state_total")))

```
# Data Analytics Term Project
## Project Desription

### Topic:
The effect of drug decriminalization in Oregon on overdose deaths and well-being compared to California and Washington.

### Why it Matters 
The study of drug decriminalization and overdose rates in Oregon has potential for significant policy and societal relevance. There are many debates on the benefits and drawbacks that drug use and decriminalization has on potential for policy implications involving economics, public health, criminal justice, and community well-being.

### Research Question and Motivation
What is the effect of drug decriminalization on overdose deaths and well-being? A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths, economic impacts on health care costs, and overall well-being. Investigating the effects with county level data, we used a  difference-in-difference model with fixed effects to assess the impact of the policy change in Oregon. By comparing the outcomes in Oregon before and after the intervention to a control group of Washington and California counties as control groups.

### Description of Data

#### Overdose Mortality by County in the US:
The Overdose Mortality by County in the US is from the CDC Center for Health Statistics for the years 2020-2023. These provisional county-Level drug Overdose death counts are based on death records and is received by the NCHS receives the death records from the state vital registration offices through the Vital Statistics Cooperative Program (VSCP). The provisional counts are reported by the county were the decedent resided, not necessarily where they died.

#### County Well-Being Rankings:
The County Well-Being Rankings are for the years 2014-2023 and is collected by County Health Rankings & Roadmaps CHR&R, a program of the University of Wisconsin Population Health Institute. The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public health rankings.

#### Overdose by Drug:
The Overdose by drug dataset contains information on non-fatal overdose data by state for 2018-2023. Provided by the CDC, DOSE (Drug Overdose Surveillance and Epidemiology) collects health department reports in each state.

Other variables of interest include opioid and stimulant overdoses and the significance of the change from previous months in those overdoses.
Violent Crime Data From the FBI crime data reporter database, the number of violent crimes in Oregon.

#### How are they related? 
The Well-Being Ranks, Overdose Mortality by County contain county fips codes, county names and year in relation. The Overdose by Drug has a relationship of total non-fatal drug overdoses in each state.

## Data Processing 
For the county health rankings data, we originally had years 2014-2023, but after further inspection we found the way overdose deaths were collected for 2014-2015 was significantly different than for the years 2016-2023. We opted to drop those years and due to those data sets not including as many well-being measures.

The data collected was generally in clean condition. Some column name changes were done, and general data manipulation was completed. For example, health rankings came from yearly filet, which had to be merged together. There was generally little missing data. Violent crime and overdose by drugs contained zero missing values. However, a key data set used, health rankings, contained 13 percent N/A values. These appeared generally randomly scatter, and we will assume these are missing at random and will be ignored.

### Findings and Visualizations

Initial findings include general decrease in well being, after 2020 which is likely due to Covid-19. Well-being variables include, mental health days, fair or poor health days, excessive drinking, and income. Additionally, drug overdose deaths have been increasing, especially since 2020. However, it appears that Oregon's increase may be slower relative to Washington and Oregon after decriminalizing drugs.

We compared the parallel trends of overdose deaths per-capita in all states. Pre-treatment shows (2016-2020) shows

``` {r}
# Show deaths per-capita for each state
data2 <- health_ranking |>
  mutate(od_deaths_percapita = combined_overdose_deaths / population)

# All states and overdose death per-capita for years 2016, 2020, 2023
data2 |>
  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
  scale_x_continuous(
    trans = "log10",
    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
  ) +
  geom_density(alpha = 0.3, position = "stack") +
  facet_grid(year ~ .) +
  labs(
    title = "Density Plot of Overdose Deaths Per Capita by State",
    x = "Overdose Deaths Per Capita",
    y = "Count of Overdose Deaths",
    fill = "State"
  )

```

Comparing the rate of drug mortality by county depicts increases in many counties in every state we are examining.

```{r Cleveland dot plots, fig.height=5, fig.width=8}

# Create Cleveland dot plots for each state to show changes in county before and after the policy change 
# Arrange the data mortality in desc order and subset into states to see the changes in county by year
health_ranking_OR <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "Oregon")

health_ranking_WA <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "Washington")

health_ranking_CA <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "California")

# Oregon
health_ranking_OR %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Oregon",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

# Washington
health_ranking_WA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Washington",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()
```

```{r CA Cleveland dot plots, fig.height=7, fig.width=8}
# California 
# Save the ggplot to a variable

health_ranking_CA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in California",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

```

### Methods

### Results
Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.

### Limitations 
It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths. Since multiple factors can contribute to drug-related outcomes. There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health. There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon drug policies and may not be generalized to other regions or policies. There is also limited data post treatment period (2 years). Additionally, the synthetic Oregon may not be a good fit. This will be further analyzed throughout the semester. As well as continuing to fully understand the statistical basis for synthetic control better.

### 2023-11-29 Code/Draft1Rmd.Rmd
Added Line Plots

@@
 library(ggplot2)
+source("Functions.R")
 
@@
 
+```{r}
+#Create individual data sets for Oregon, Washington, and California
+or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
+ca_health_ranking <- health_ranking[health_ranking$state == "California",]
+wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]
+
+#Now create yearly averages for the states from the healthrank_to_yearly function from Functions.R:
+or_yearly <- healthrank_to_yearly(or_health_ranking)
+ca_yearly <- healthrank_to_yearly(ca_health_ranking)
+wa_yearly <- healthrank_to_yearly(wa_health_ranking)
+
+ggplot()+
+  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths, color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths, color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
+  scale_color_manual(name = "State",
+                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  labs(y = "Overdose Deaths") +
+  theme_bw()
+
+#Now do a log transformation to better show percent differences due to population size differences
+ggplot()+
+  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
+  scale_color_manual(name = "State",
+                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  theme_bw()
+```
+
 Comparing the rate of drug mortality by county depicts increases in many counties in every state we are examining.

### 2023-11-29 Code/Draft1Rmd.tex
Added Line Plots

--- 2023-11-29 Code/Draft1Rmd.Rmd
@@
----
-title: "Exploratory Analysis"
-author: "James Santarpio and Mardi Elings"
-date: "2023-11-30"
-output:
-  pdf_document: default
-  html_document:
-    df_print: paged
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math} % this also loads fontspec
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Exploratory Analysis},
+  pdfauthor={James Santarpio and Mardi Elings},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+\title{Exploratory Analysis}
+\author{James Santarpio and Mardi Elings}
+\date{2023-11-30}
 
-library(tidyverse)
-library(ggplot2)
+\begin{document}
+\maketitle
 
-url <- c("https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Cleaned%20Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
-# Read csv
-data <- read.csv(url)
+\hypertarget{data-analytics-term-project}{%
+\section{Data Analytics Term
+Project}\label{data-analytics-term-project}}
 
-# Remove 2023- 2023 is a duplicate of 2022 for much of the data
-health_ranking <- data %>%
-  filter(year <= 2023) %>%
-  filter(! (county %in% c("year_state_total")))
+\hypertarget{project-desription}{%
+\subsection{Project Desription}\label{project-desription}}
 
-```
-# Data Analytics Term Project
-## Project Desription
+\hypertarget{topic}{%
+\subsubsection{Topic:}\label{topic}}
 
-### Topic:
-The effect of drug decriminalization in Oregon on overdose deaths and well-being compared to California and Washington.
+The effect of drug decriminalization in Oregon on overdose deaths and
+well-being compared to California and Washington.
 
-### Why it Matters 
-The study of drug decriminalization and overdose rates in Oregon has potential for significant policy and societal relevance.
-There are many debates on the benefits and drawbacks that drug use and decriminalization has on potential for policy implications involving economics, public health, criminal justice, and community well-being.
+\hypertarget{why-it-matters}{%
+\subsubsection{Why it Matters}\label{why-it-matters}}
 
-### Research Question and Motivation
-What is the effect of drug decriminalization on overdose deaths and well-being?
-A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths, economic impacts on health care costs, and overall well-being.
-Investigating the effects with county level data, we used a  difference-in-difference model with fixed effects to assess the impact of the policy change in Oregon.
-By comparing the outcomes in Oregon before and after the intervention to a control group of Washington and California counties as control groups.
+The study of drug decriminalization and overdose rates in Oregon has
+potential for significant policy and societal relevance. There are many
+debates on the benefits and drawbacks that drug use and
+decriminalization has on potential for policy implications involving
+economics, public health, criminal justice, and community well-being.
 
-### Description of Data
+\hypertarget{research-question-and-motivation}{%
+\subsubsection{Research Question and
+Motivation}\label{research-question-and-motivation}}
 
-#### Overdose Mortality by County in the US:
-The Overdose Mortality by County in the US is from the CDC Center for Health Statistics for the years 2020-2023.
-These provisional county-Level drug Overdose death counts are based on death records and is received by the NCHS receives the death records from the state vital registration offices through the Vital Statistics Cooperative Program (VSCP).
-The provisional counts are reported by the county were the decedent resided, not necessarily where they died.
+What is the effect of drug decriminalization on overdose deaths and
+well-being? A policy change like Oregon's can have significant
+consequences on populations who are most at risk of overdose deaths,
+economic impacts on health care costs, and overall well-being.
+Investigating the effects with county level data, we used a
+difference-in-difference model with fixed effects to assess the impact
+of the policy change in Oregon. By comparing the outcomes in Oregon
+before and after the intervention to a control group of Washington and
+California counties as control groups.
 
-#### County Well-Being Rankings:
-The County Well-Being Rankings are for the years 2014-2023 and is collected by County Health Rankings & Roadmaps CHR&R, a program of the University of Wisconsin Population Health Institute.
-The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public health rankings.
+\hypertarget{description-of-data}{%
+\subsubsection{Description of Data}\label{description-of-data}}
 
-#### Overdose by Drug:
-The Overdose by drug dataset contains information on non-fatal overdose data by state for 2018-2023.
-Provided by the CDC, DOSE (Drug Overdose Surveillance and Epidemiology) collects health department reports in each state.
+\hypertarget{overdose-mortality-by-county-in-the-us}{%
+\paragraph{Overdose Mortality by County in the
+US:}\label{overdose-mortality-by-county-in-the-us}}
 
-Other variables of interest include opioid and stimulant overdoses and the significance of the change from previous months in those overdoses.
-Violent Crime Data From the FBI crime data reporter database, the number of violent crimes in Oregon.
+The Overdose Mortality by County in the US is from the CDC Center for
+Health Statistics for the years 2020-2023. These provisional
+county-Level drug Overdose death counts are based on death records and
+is received by the NCHS receives the death records from the state vital
+registration offices through the Vital Statistics Cooperative Program
+(VSCP). The provisional counts are reported by the county were the
+decedent resided, not necessarily where they died.
 
-#### How are they related? 
-The Well-Being Ranks, Overdose Mortality by County contain county fips codes, county names and year in relation.
-The Overdose by Drug has a relationship of total non-fatal drug overdoses in each state.
+\hypertarget{county-well-being-rankings}{%
+\paragraph{County Well-Being
+Rankings:}\label{county-well-being-rankings}}
 
-## Data Processing 
-For the county health rankings data, we originally had years 2014-2023, but after further inspection we found the way overdose deaths were collected for 2014-2015 was significantly different than for the years 2016-2023.
-We opted to drop those years and due to those data sets not including as many well-being measures.
+The County Well-Being Rankings are for the years 2014-2023 and is
+collected by County Health Rankings \& Roadmaps CHR\&R, a program of the
+University of Wisconsin Population Health Institute. The health rankings
+include deaths which include overdose deaths, years of potential life
+lost, mental health statistics, income, and other health public health
+rankings.
 
-The data collected was generally in clean condition.
-Some column name changes were done, and general data manipulation was completed.
-For example, health rankings came from yearly filet, which had to be merged together.
-There was generally little missing data.
-Violent crime and overdose by drugs contained zero missing values.
-However, a key data set used, health rankings, contained 13 percent N/A values.
-These appeared generally randomly scatter, and we will assume these are missing at random and will be ignored.
+\hypertarget{overdose-by-drug}{%
+\paragraph{Overdose by Drug:}\label{overdose-by-drug}}
 
-### Findings and Visualizations
+The Overdose by drug dataset contains information on non-fatal overdose
+data by state for 2018-2023. Provided by the CDC, DOSE (Drug Overdose
+Surveillance and Epidemiology) collects health department reports in
+each state.
 
-Initial findings include general decrease in well being, after 2020 which is likely due to Covid-19.
-Well-being variables include, mental health days, fair or poor health days, excessive drinking, and income.
-Additionally, drug overdose deaths have been increasing, especially since 2020.
-However, it appears that Oregon's increase may be slower relative to Washington and Oregon after decriminalizing drugs.
+Other variables of interest include opioid and stimulant overdoses and
+the significance of the change from previous months in those overdoses.
+Violent Crime Data From the FBI crime data reporter database, the number
+of violent crimes in Oregon.
 
-We compared the parallel trends of overdose deaths per-capita in all states. Pre-treatment shows (2016-2020) shows
+\hypertarget{how-are-they-related}{%
+\paragraph{How are they related?}\label{how-are-they-related}}
 
-``` {r}
-# Show deaths per-capita for each state
-data2 <- health_ranking |>
-  mutate(od_deaths_percapita = combined_overdose_deaths / population)
+The Well-Being Ranks, Overdose Mortality by County contain county fips
+codes, county names and year in relation. The Overdose by Drug has a
+relationship of total non-fatal drug overdoses in each state.
 
-# All states and overdose death per-capita for years 2016, 2020, 2023
-data2 |>
-  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
-  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
-  scale_x_continuous(
-    trans = "log10",
-    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
-  ) +
-  geom_density(alpha = 0.3, position = "stack") +
-  facet_grid(year ~ .) +
-  labs(
-    title = "Density Plot of Overdose Deaths Per Capita by State",
-    x = "Overdose Deaths Per Capita",
-    y = "Count of Overdose Deaths",
-    fill = "State"
-  )
+\hypertarget{data-processing}{%
+\subsection{Data Processing}\label{data-processing}}
 
-```
+For the county health rankings data, we originally had years 2014-2023,
+but after further inspection we found the way overdose deaths were
+collected for 2014-2015 was significantly different than for the years
+2016-2023. We opted to drop those years and due to those data sets not
+including as many well-being measures.
 
-Comparing the rate of drug mortality by county depicts increases in many counties in every state we are examining.
+The data collected was generally in clean condition. Some column name
+changes were done, and general data manipulation was completed. For
+example, health rankings came from yearly filet, which had to be merged
+together. There was generally little missing data. Violent crime and
+overdose by drugs contained zero missing values. However, a key data set
+used, health rankings, contained 13 percent N/A values. These appeared
+generally randomly scatter, and we will assume these are missing at
+random and will be ignored.
 
-```{r Cleveland dot plots, fig.height=5, fig.width=8}
+\hypertarget{findings-and-visualizations}{%
+\subsubsection{Findings and
+Visualizations}\label{findings-and-visualizations}}
 
-# Create Cleveland dot plots for each state to show changes in county before and after the policy change 
-# Arrange the data mortality in desc order and subset into states to see the changes in county by year
-health_ranking_OR <- health_ranking |>
-  arrange(desc(combined_overdose_mortality_rate)) |>
-  filter(state == "Oregon")
+Initial findings include general decrease in well being, after 2020
+which is likely due to Covid-19. Well-being variables include, mental
+health days, fair or poor health days, excessive drinking, and income.
+Additionally, drug overdose deaths have been increasing, especially
+since 2020. However, it appears that Oregon's increase may be slower
+relative to Washington and Oregon after decriminalizing drugs.
 
-health_ranking_WA <- health_ranking |>
-  arrange(desc(combined_overdose_mortality_rate)) |>
-  filter(state == "Washington")
+We compared the parallel trends of overdose deaths per-capita in all
+states. Pre-treatment shows (2016-2020) shows
 
-health_ranking_CA <- health_ranking |>
-  arrange(desc(combined_overdose_mortality_rate)) |>
-  filter(state == "California")
+\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-1-1.pdf}
 
-# Oregon
-health_ranking_OR %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
-  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
-  ggplot(aes(combined_overdose_mortality_rate, county)) +
-  geom_line(aes(group = county)) +
-  geom_point(size = 2, aes(color = factor(year))) +
-  labs(title = "Overdose Mortality Rate in Oregon",
-       x = "Overdose Mortality Rate",  # Add x-axis label
-       y = "County") +
-  theme_minimal()
+\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-2-1.pdf}
+\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-2-2.pdf}
 
-# Washington
-health_ranking_WA %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
-  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
-  ggplot(aes(combined_overdose_mortality_rate, county)) +
-  geom_line(aes(group = county)) +
-  geom_point(size = 2, aes(color = factor(year))) +
-  labs(title = "Overdose Mortality Rate in Washington",
-       x = "Overdose Mortality Rate",  # Add x-axis label
-       y = "County") +
-  theme_minimal()
-```
+Comparing the rate of drug mortality by county depicts increases in many
+counties in every state we are examining.
 
-```{r CA Cleveland dot plots, fig.height=7, fig.width=8}
-# California 
-# Save the ggplot to a variable
+\includegraphics{Draft1Rmd_files/figure-latex/Cleveland dot plots-1.pdf}
+\includegraphics{Draft1Rmd_files/figure-latex/Cleveland dot plots-2.pdf}
 
-health_ranking_CA %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
-  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
-  ggplot(aes(combined_overdose_mortality_rate, county)) +
-  geom_line(aes(group = county)) +
-  geom_point(size = 2, aes(color = factor(year))) +
-  labs(title = "Overdose Mortality Rate in California",
-       x = "Overdose Mortality Rate",  # Add x-axis label
-       y = "County") +
-  theme_minimal()
+\includegraphics{Draft1Rmd_files/figure-latex/CA Cleveland dot plots-1.pdf}
 
-```
+\hypertarget{methods}{%
+\subsubsection{Methods}\label{methods}}
 
-### Methods
+\hypertarget{results}{%
+\subsubsection{Results}\label{results}}
 
-### Results
-Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.
+Decriminalizing drugs in Oregon has slowed the increase in drug overdose
+deaths.
 
-### Limitations 
-It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths.
-Since multiple factors can contribute to drug-related outcomes.
-There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health.
-There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon drug policies and may not be generalized to other regions or policies.
-There is also limited data post treatment period (2 years).
-Additionally, the synthetic Oregon may not be a good fit.
-This will be further analyzed throughout the semester.
-As well as continuing to fully understand the statistical basis for synthetic control better.
+\hypertarget{limitations}{%
+\subsubsection{Limitations}\label{limitations}}
+
+It is a challenge to find a single cause for an impact on drug
+decriminalization and its relationship to drug overdose deaths. Since
+multiple factors can contribute to drug-related outcomes. There has been
+some evidence linked to the Covid-19 pandemic contributing to drug use,
+well-being, and mental health. There is limited generalizability to this
+study as findings from synthetic control analysis are specific to the
+context of Oregon drug policies and may not be generalized to other
+regions or policies. There is also limited data post treatment period (2
+years). Additionally, the synthetic Oregon may not be a good fit. This
+will be further analyzed throughout the semester. As well as continuing
+to fully understand the statistical basis for synthetic control better.
+
+\end{document}

### 2023-11-30 Code/Draft1Rmd.Rmd
jsjsjs
Updated line plot

--- 2023-11-29 Code/Draft1Rmd.Rmd
@@
 
+```{r}
+#Create individual data sets for Oregon, Washington, and California
+or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
+ca_health_ranking <- health_ranking[health_ranking$state == "California",]
+wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]
+
+#Now create yearly averages for the states from the healthrank_to_yearly function from Functions.R:
+or_yearly <- healthrank_to_yearly(or_health_ranking)
+ca_yearly <- healthrank_to_yearly(ca_health_ranking)
+wa_yearly <- healthrank_to_yearly(wa_health_ranking)
+
+ggplot()+
+  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
+  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
+  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
+  scale_color_manual(name = "State",
+                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  theme_bw()
+```
+
 Comparing the rate of drug mortality by county depicts increases in many counties in every state we are examining.

### 2023-11-30 Code/Draft1Rmd.Rmd
mehtods update Draft1Rmd.Rmd

@@
 library(ggplot2)
-source("Functions.R")
+library(Synth)
 
@@
 health_ranking <- data %>%
-  filter(year <= 2023) %>%
+  filter(year <= 2022) %>%
   filter(! (county %in% c("year_state_total")))
@@
 
+``` {r}
+# Show deaths per-capita for each state
+data2 <- health_ranking |>
+  mutate(od_deaths_percapita = combined_overdose_deaths / population)
+
+# All states and overdose death per-capita for years 2016, 2020, 2023
+data2 |>
+  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
+  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
+  scale_x_continuous(
+    trans = "log10",
+    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
+  ) +
+  geom_density(alpha = 0.3, position = "stack") +
+  facet_grid(year ~ .) +
+  labs(
+    title = "Density Plot of Overdose Deaths Per Capita by State",
+    x = "Overdose Deaths Per Capita",
+    y = "Count of Overdose Deaths",
+    fill = "State"
+  )
+
+```
+
 ```{r}
@@
 
-Comparing the rate of drug mortality by county depicts increases in many counties in every state we are examining.
+Comparing the rate of drug mortality by county depicts increases in many counties in all three states we are examining.
 
@@
 
-### Methods
+## Methods
+
+## Difference-in-Difference 
+We used a difference-in-difference regression with fixed effects to investigate the impact of drug decriminalization in Oregon counties compared to other states.
+
+```{r results="hide" }
+# Create a DiD variable
+did_data <- health_ranking |>
+  mutate(treatment = ifelse(state == "Oregon" & year >= 2021, 1, 0))
+
+# Fit the DiD model using year and county fixed effects and population as a covariate 
+did_model_county <- lm(combined_overdose_deaths ~ treatment + factor(county) + factor(year) + treatment:factor(year) + population, data = did_data)
+summary(did_model_county)
+
+```
+
+### Synthetic Control 
+```{r results="hide"}
+#Synthetic Control
+
+#Create indivdual data sets for Oregon, Washington, and California
+or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
+ca_health_ranking <- health_ranking[health_ranking$state == "California",]
+wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]
+
+healthrank_to_yearly <- function(df){
+  df <- df[!df$county == "year_state_total",]
+  df <- df %>%
+    group_by(year) %>%
+    summarise(across(where(is.numeric), ~ mean(., na.rm = TRUE))) %>%
+    select(-c("X", "fips"))
+  return(df)
+}
+
+#Now create yearly averages for the states from the healthrank_to_yearly function from Functions.R:
+or_yearly <- healthrank_to_yearly(or_health_ranking)
+ca_yearly <- healthrank_to_yearly(ca_health_ranking)
+wa_yearly <- healthrank_to_yearly(wa_health_ranking)
+
+#The three states follow trends mostly closely (Relative to each scale)- Good fit for synthetic control
+
+#Create state number variables to use for synthetic control
+health_ranking$state_num <- ifelse(health_ranking$state == "Oregon", 1, 2)
+
+health_ranking$state_num <- ifelse(health_ranking$state == "Oregon", 1,
+                                   ifelse(health_ranking$state == "California", 2, 3))
+
+#Lets create long data set for each year- for the synth package
+#Add the state numbers and states to the yearly data sets
+or_yearly$state_num <- 1
+or_yearly$X.1 <- "Oregon"
+ca_yearly$state_num <- 2
+ca_yearly$X.1 <- "California"
+wa_yearly$state_num <- 3
+wa_yearly$X.1 <- "Washington"
+
+#Join the three yearly data sets by row
+health_ranking_yearly <- rbind(rbind(or_yearly, ca_yearly), wa_yearly)
+health_ranking_yearly <- as.data.frame(health_ranking_yearly) #for synth package purposes
+
+#Run a synthetic control with the predictor variable being combined overdose deaths. The two states Washington and 
+#California will serve as the donor pool states. Variables used will be smokers, percent fair or poor health, excessive drinking, and unemployed.
+#The training data will be from 2016-2020, with 2021 and 2022 being the treatment outcomes section.
+dataprep.out <-
+  dataprep(health_ranking_yearly,
+           predictors = c("smokers", "percent_fair_or_poor_health", "excessive_drinking"),
+           dependent = "combined_overdose_deaths",
+           unit.variable = "state_num",
+           time.variable = "year",
+           unit.names.variable = "X.1",
+           treatment.identifier = 1,
+           controls.identifier = c(2,3),
+           time.predictors.prior = c(2016:2020),
+           time.optimize.ssr = c(2016:2020),
+           time.plot = 2016:2022,
+  )
+
+synth.out <- synth(dataprep.out)
+print(synth.tables <- synth.tab(
+  dataprep.res = dataprep.out,
+  synth.res = synth.out)
+)
+
+#Create a plot of pre and post treatment
+path.plot(synth.res = synth.out,
+          dataprep.res = dataprep.out,
+          Ylab = c("Y"),
+          Xlab = c("Year"),
+          Legend = c("Oregon", "Synthetic Oregon"),
+          Legend.position = c("topleft")
+)
+
+```
 

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated line plot

@@
 library(Synth)
+source("Functions.R")
 
@@
                      values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+  labs(y = "Log Combined Overdose Deaths")
   theme_bw()
@@
 
+```
+
+```{r results="hide"}
+or_yearly2 <- or_yearly
+or_yearly2$
 ```

### 2023-11-30 Code/Functions.R

@@ -17,7 +17,7 @@ healthrank_to_yearly <- function(df){
   df <- df[!df$county == "year_state_total",]
   df <- df %>%
     group_by(year) %>%
-    summarise(across(where(is.numeric), ~ mean(., na.rm = TRUE))) %>%
+    summarise(across(where(is.numeric), ~ sum(., na.rm = TRUE))) %>%
     select(-c("X", "fips"))
   return(df)
 }

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated results

@@
 
+A synthetic control analysis was conducted using Washington and Oregon as donor states.
+This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.
 ## Difference-in-Difference 
@@
 
-```
-
-```{r results="hide"}
-or_yearly2 <- or_yearly
-or_yearly2$
 ```
@@
 ### Results
+
+Due to the limitations surrounding our synthetic control analysis, our synthetic Oregon was higher than the actual Oregon pre treatment which should not have happened.
+This may be due to a lack of accurate covariates used.
+Our analysis shows synthetic Oregon increasing more than actual Oregon, however limitations have to be noted surrounding this conclusion.
+
+Diff in Diff:
+
 Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated line plot

@@
 
-ggplot()+
+combined_yearly <- healthrank_to_yearly(health_ranking)
+
+ggplot() +
   geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
@@
   geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
+  geom_point(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon"), size = 1) +
+  geom_point(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California"), size = 1) +
+  geom_point(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington"), size = 1) +
   scale_color_manual(name = "State",
                      values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
-  labs(y = "Log Combined Overdose Deaths")
+  labs(y = "Log Combined Overdose Deaths") +
   theme_bw()
+```
 ```
@@
 This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.
+
 ## Difference-in-Difference 

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated limitation

@@
   )
-
 synth.out <- synth(dataprep.out)
@@
 Additionally, the synthetic Oregon may not be a good fit.
-This will be further analyzed throughout the semester.
-As well as continuing to fully understand the statistical basis for synthetic control better.

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated limitations

@@
 ### Limitations 
+Oregon drug decriminilization has recently went into effect. The data for overdose deaths after the law has been enacted is limited with only 2021 and 2022. 
 It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths.

### 2023-11-30 Code/Draft1Rmd.Rmd
Updated line plot

@@
 
-combined_yearly <- healthrank_to_yearly(health_ranking)
-
 ggplot() +
@@
   geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
-  geom_point(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon"), size = 1) +
-  geom_point(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California"), size = 1) +
-  geom_point(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington"), size = 1) +
+  geom_point(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon"), size = 3) +
+  geom_point(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California"), size = 3) +
+  geom_point(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington"), size = 3) +
+  geom_vline(xintercept = as.numeric("2021"), linetype = "dashed", color = "black", size = 1) +
   scale_color_manual(name = "State",

### 2023-11-30 Code/Draft1Rmd.Rmd
Update Draft1Rmd.Rmd

@@
 
-A synthetic control analysis was conducted using Washington and Oregon as donor states.
-This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.
-
 ## Difference-in-Difference 
-We used a difference-in-difference regression with fixed effects to investigate the impact of drug decriminalization in Oregon counties compared to other states.
+We used a difference-in-difference regression with fixed effects to investigate the impact of drug decriminalization starting in 2021 in Oregon counties compared to other states.
+The treatment effect shows a 80.39 decrease in overdose deaths compared to the control group and is statistically significant (p < 0.001).
+The fixed effects for years 2019 onwards are statistically signifcant, which indicate a time trend.
 
@@
 ### Synthetic Control 
+A synthetic control analysis was conducted using Washington and Oregon as donor states.
+This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.
+
 ```{r results="hide"}

### 2023-11-30 Code/Draft1Rmd.Rmd
update density plot Draft1Rmd.Rmd

@@
 
-url <- c("https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Cleaned%20Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
+# Replace 'your/path/to/directory' with the actual path to your desired working directory
+setwd("your/path/to/directory")
+
+# Change file path name to your directory
+file_path <- c("project/cleaned/MergedAllCountyRanks2.csv")
 # Read csv
-data <- read.csv(url)
+data <- read.csv(file_path)
 
@@
 
-``` {r}
+``` {r warnings = FALSE}
 # Show deaths per-capita for each state
@@
   ) +
-  geom_density(alpha = 0.3, position = "stack") +
+  geom_density(alpha = 0.3, position = "stack", linetype = "solid", size = 1) +  # Use linewidth instead of size
   facet_grid(year ~ .) +
@@
     fill = "State"
-  )
+  ) +
+  theme_minimal()  # Customize the theme if needed
 

### 2023-11-30 Code/Draft1Rmd.Rmd
knit warnings upated Draft1Rmd.Rmd
density description Draft1Rmd.Rmd

--- 2023-11-30 Code/Draft1Rmd.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
 
@@
 
-We compared the parallel trends of overdose deaths per-capita in all states. Pre-treatment shows (2016-2020) shows
+We compared the parallel trends of overdose deaths per-capita in all states using stacked kernel density plots.
+Pre-treatment shows (2016-2020) fairly similar trends, especially in Oregon and Washington.
+California has the highest proportion of overdoses in all years.
 
-``` {r}
+``` {r warnings = FALSE, fig.height=5, fig.width= 10 }
 # Show deaths per-capita for each state
@@
 
-# All states and overdose death per-capita for years 2016, 2020, 2023
 data2 |>
-  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
+  filter(year %in% c(2016, 2018, 2020, 2022) & !is.na(od_deaths_percapita)) |>
   ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
@@
   ) +
-  geom_density(alpha = 0.3, position = "stack") +
+  geom_density(alpha = 0.3, position = "stack", linetype = "solid", linewidth = 1) +  # Use linewidth instead of size
   facet_grid(year ~ .) +

### 2023-11-30 Code/Draft1Rmd.Rmd
did  update

@@
 library(Synth)
+library(knitr)
 source("Functions.R")
@@
 
-``` {r warnings = FALSE, fig.height=5, fig.width= 10 }
+``` {r warnings = FALSE, fig.height=6, fig.width= 10 }
 # Show deaths per-capita for each state
@@
 
-``` {r}
-# Show deaths per-capita for each state
-data2 <- health_ranking |>
-  mutate(od_deaths_percapita = combined_overdose_deaths / population)
+There is an increase in overdose deaths in all states after 2020, which is most likely due to the Covid-19 pandemic.
+The policy intervention was implemented in Oregon in 2021, and while the drug overdose rates are still increasing, Oregon appears to be increasing at a slightly slower rate than Washington and Oregon.
 
-# All states and overdose death per-capita for years 2016, 2020, 2023
-data2 |>
-  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
-  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
-  scale_x_continuous(
-    trans = "log10",
-    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
-  ) +
-  geom_density(alpha = 0.3, position = "stack") +
-  facet_grid(year ~ .) +
-  labs(
-    title = "Density Plot of Overdose Deaths Per Capita by State",
-    x = "Overdose Deaths Per Capita",
-    y = "Count of Overdose Deaths",
-    fill = "State"
-  )
-
-```
-
-```{r}
+```{r fig.height=5, fig.width= 10}
 #Create individual data sets for Oregon, Washington, and California
@@
   scale_color_manual(name = "State",
-                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
+                     values = c("Oregon" = "green", "California" = "red", "Washington" = "blue")) +
   labs(y = "Log Combined Overdose Deaths") +
@@
 ```
-```
 
-Comparing the rate of drug mortality by county depicts increases in many counties in all three states we are examining.
+Comparing the rate of drug mortality by county depicts increases after 2020 in many counties with noticeable increases in 2022 for all three states we are examining.
 
@@
 health_ranking_OR %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
+  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
   mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
@@
 health_ranking_WA %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
+  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
   mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
@@
 # California
-# Save the ggplot to a variable
 
 health_ranking_CA %>%
-  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
+  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
   mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
@@
 
-## Methods
+## Methods and Results
 
-## Difference-in-Difference 
-We used a difference-in-difference regression with fixed effects to investigate the impact of drug decriminalization starting in 2021 in Oregon counties compared to other states.
-The treatment effect shows a 80.39 decrease in overdose deaths compared to the control group and is statistically significant (p < 0.001).
-The fixed effects for years 2019 onwards are statistically signifcant, which indicate a time trend.
+### Difference-in-Difference 
+We used a difference-in-difference regression with year and county fixed effects to investigate the impact of drug decriminalization starting in 2021 in Oregon counties compared to Washington and California.
+The treatment estimates  an average effect of a 80.39 decrease in overdose deaths compared to the control group and is statistically significant (p < 0.001).
+The fixed effects for years 2019 onwards are statistically significant, which indicate a trend over time.
+County specific fixed effects are included in the model, represented by the coefficients for each county.
+Counties like Los Angeles and Riverside have positive coefficients, indicating higher overdose deaths compared to the baseline county.
 
@@
 
+```{r results='asis'}
+library(tidyverse)
+library(broom)
+library(huxtable)
+library(jtools)
+# Get coefficients, standard errors, and p-values
+coefficients_to_display <- coef(summary(did_model_county))[c("(Intercept)", "treatment", "population"), c("Estimate", "Std. Error", "Pr(>|t|)")]
+
+# Create a data frame for the coefficients
+coefficients_df <- as.data.frame(coefficients_to_display)
+
+# Round coefficients to 2 decimal places
+rounded_coefficients <- round(coefficients_df, 3)
+
+# Increase decimals for p-values
+rounded_coefficients$`Pr(>|t|)` <- format(rounded_coefficients$`Pr(>|t|)`, nsmall = 6)  # Increase to desired number of decimals
+
+# Format large numbers with commas
+formatted_coefficients <- format(rounded_coefficients, big.mark = ",")
+
+# Display the formatted coefficients using kable with additional formatting
+kable(
+  formatted_coefficients,
+  format = "markdown",
+  align = c("l", "c", "c", "c"),  # Adjust alignment as needed
+  col.names = c("Variable", "Coefficient", "Std. Error", "P-value"),  # Rename column headers
+  caption = "Regression Coefficients",  # Add a caption
+  escape = FALSE  # Disable HTML escaping for formatting
+)
+
+```
+
 ### Synthetic Control 
-A synthetic control analysis was conducted using Washington and Oregon as donor states.
-This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.
+A synthetic control analysis was conducted using Washington and California as donor states.
+This analysis has significant limitations due to the lack of donor states and potential lack of credible covariates.However, it does show that the synthetic "Oregon" with more overdose deaths if the policy intervention did not exist.
 
@@
 
-### Results
-
-Due to the limitations surrounding our synthetic control analysis, our synthetic Oregon was higher than the actual Oregon pre treatment which should not have happened.
-This may be due to a lack of accurate covariates used.
-Our analysis shows synthetic Oregon increasing more than actual Oregon, however limitations have to be noted surrounding this conclusion.
-
-Diff in Diff:
-
-Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.
-
 ### Limitations 
-Oregon drug decriminilization has recently went into effect. The data for overdose deaths after the law has been enacted is limited with only 2021 and 2022. 
+Oregon drug decriminalization has just recently gone into effect.
+The data for overdose deaths after the law has been enacted is limited with only 2021 and 2022.
 It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths.
@@
 There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health.
-There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon drug policies and may not be generalized to other regions or policies.
+
 There is also limited data post treatment period (2 years).
-Additionally, the synthetic Oregon may not be a good fit.
+Difference-in-Differences (DiD) estimates, particularly with fixed effects, are context-specific and may lack generalizability beyond the studied time period and treatment group.The validity of DiD estimates relies on the assumption of parallel trends in the absence of treatment.
+While fixed effects enhance internal validity, they may limit generalizability to settings with different characteristics.
+
+Similarly, the synthetic control analysis is specific to the context of Oregon drug policies and may not be generalized to other regions or policies.
+There is also limited data post treatment period (2 years).
+Further analysis could be done using all 50 states to find a model that best fits a synthetic Oregon.

### 2023-11-30 Code/visualizations.R
added plots

@@ -1,9 +1,9 @@
 library(tidyverse)
 
 # Load from Github
-url <- c("https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Cleaned%20Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
+file_path <- c("cleaned data/merged_health_rankings")
 # Read csv
-data <- read.csv(url)
+data <- read.csv(file_path)
 # Remove Key
 data <- data |>
   select(-X, -X.1)
@@ -16,11 +16,9 @@ ggplot(data, aes(x = combined_overdose_mortality_rate)) +
   geom_histogram(binwidth = 1, fill = "blue") +
   labs(title = "Distribution of Overdose Mortality Rate")
 
-
 ggplot(data, aes(x = state)) +
   geom_bar(fill = "green") +
   labs(title = "State Distribution")
-
 
 ggplot(data, aes(x = combined_overdose_mortality_rate, y = county)) +
   geom_point(color = "red") +
@@ -38,41 +36,65 @@ dataCA <- data |>
   filter(state == "California")
 
 # Create a cleveland dot plot for each state to show changes in county before and after the policy change 
+
 # Oregon
-dataOR |>
-  filter(year %in% c(2020, 2023) & !is.na(combined_overdose_mortality_rate)) |>
-  mutate(county = reorder(county, combined_overdose_mortality_rate)) |>
+health_ranking_OR %>%
+  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
+  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
   ggplot(aes(combined_overdose_mortality_rate, county)) +
   geom_line(aes(group = county)) +
-  geom_point(size=2, aes(color = factor(year)))
+  geom_point(size = 2, aes(color = factor(year))) +
+  labs(title = "Overdose Mortality Rate in Oregon",
+       x = "Overdose Mortality Rate",  # Add x-axis label
+       y = "County") +
+  theme_minimal()
 
 # Washington
-dataWA |>
-  filter(year %in% c(2020, 2023) & !is.na(combined_overdose_mortality_rate)) |>
-  mutate(county = reorder(county, combined_overdose_mortality_rate)) |>
+health_ranking_WA %>%
+  filter(year %in% c(2016, 2020, 2022) & !is.na(combined_overdose_mortality_rate)) %>%
[…]

### 2023-11-30 Code/Main.R
added table Main.R

@@ -6,6 +6,9 @@ source("Code/Functions.R")
 
 library(tidyverse)
 library(ggplot2)
+library(broom)
+library(huxtable)
+library(jtools)
 
 health_ranking <- read.csv("Data/Cleaned Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
 
@@ -232,3 +235,29 @@ did_data <- health_ranking |>
 # Fit the DiD model using year and county fixed effects and population as a covariate
 did_model_county <- lm(combined_overdose_deaths ~ treatment + factor(county) + factor(year) + treatment:factor(year) + population, data = did_data)
 summary(did_model_county)
+
+#******Make a Table to Show DiD output
+# Get coefficients, standard errors, and p-values
+coefficients_to_display <- coef(summary(did_model_county))[c("(Intercept)", "treatment", "population"), c("Estimate", "Std. Error", "Pr(>|t|)")]
+
+# Create a data frame for the coefficients
+coefficients_df <- as.data.frame(coefficients_to_display)
+
+# Round coefficients to 2 decimal places
+rounded_coefficients <- round(coefficients_df, 3)
+
+# Increase decimals for p-values
+rounded_coefficients$`Pr(>|t|)` <- format(rounded_coefficients$`Pr(>|t|)`, nsmall = 6)  # Increase to desired number of decimals
+
+# Format large numbers with commas
+formatted_coefficients <- format(rounded_coefficients, big.mark = ",")
+
+# Display the formatted coefficients using kable with additional formatting
+kable(
+  formatted_coefficients,
+  format = "markdown",
+  align = c("l", "c", "c", "c"),  # Adjust alignment as needed
+  col.names = c("Variable", "Coefficient", "Std. Error", "P-value"),  # Rename column headers
+  caption = "Regression Coefficients",  # Add a caption
+  escape = FALSE  # Disable HTML escaping for formatting
+)
